---
title: "Slow–Fast Hybrid Model: 0/1 Curie–Weiss Symptoms & Precarity"
author: "Kyuri Park"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
  cache: true
---

```{r}
#| label: setup
#| include: true

## Core libraries
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(ggplot2)

## Parallelisation
library(future)
library(furrr)

## For graphical abstract layout
library(patchwork)

theme_set(theme_minimal(base_size = 13))

## Logistic helper
inv_logit <- function(x) 1/(1 + exp(-x))

## Parallel backend
plan(multisession, workers = max(1, parallel::detectCores() - 1))
```

---

# 1. Motivation

Mental health symptoms fluctuate on fast time scales (hours–days), while
socio-economic pressures such as financial precarity, discrimination, or
workload change more slowly (months–years). Standard symptom-network models:

* treat symptoms as a *closed* system,
* interpret group differences in network structure as *internal differences*.

Here we explicitly couple:

* a *fast 0/1 Curie–Weiss symptom layer* (binary symptoms with mean-field interaction),
* a *slow precarity variable* ($P_t$) that drifts gradually,
* *feedback* from symptoms back into precarity.

We show how **group differences in symptom trajectories** can arise purely from:

* different **precarity baselines** and
* different **feedback strengths**,

with **identical** fast-layer parameters.

We examine:

* **Scenario A**: Weakly nonlinear fast layer (no bistability).
* **Scenario B**: Strongly nonlinear fast layer with bistability; precarity almost fixed around different baselines.
* **Scenario C1**: Same bistable fast layer + strong two-way feedback → catastrophic drift into high-symptom regime.
* **Scenario C2**: Same bistable fast layer + moderate feedback + noise → episodic flipping in and out of high-symptom regime.

---

# 2. Fast layer: 0/1 Curie–Weiss symptoms

We use a Curie–Weiss-type model with spins

$$
s_i \in {0, 1}.
$$

The mean symptom activation is

$$
m = \frac{1}{N}\sum_i s_i \in [0,1].
$$

The local “field” for a node is

$$
H = J(m - \tfrac{1}{2}) + h_0 + \gamma_P P.
$$

We then update spins with a logistic rule:

$$
\Pr(s_i = 1 \mid m, P) = \sigma\big(\beta H\big)
= \frac{1}{1 + e^{-\beta H}}.
$$

For large enough $\beta J$ (specifically $\beta J > 4$) this mean-field system
is **bistable** for an intermediate range of $P$.

```{r}
#| label: fast-layer
#| include: true

# One fast update block for 0/1 Curie–Weiss Ising layer
# s in {0,1}; P is current precarity
ising_fast_step01 <- function(
  P,
  s,
  beta   = 1.0,
  J      = 1.0,
  h0     = 0.0,
  gammaP = 0.5,
  sweeps = 30
){
  n <- length(s)  # spins in {0,1}
  
  for (sw in seq_len(sweeps)) {
    m <- mean(s)  # mean activation in [0,1]
    
    # field for this mean-field model
    H <- J * (m - 0.5) + h0 + gammaP * P
    p <- inv_logit(beta * H)
    
    # update spins
    for (u in seq_len(n)) {
      i    <- sample.int(n, 1)
      s[i] <- rbinom(1, 1, p)
    }
  }
  s
}
```

---

# 3. Slow layer: precarity dynamics

The slow variable $P_t$ evolves with:

* **mean reversion** to group-specific baseline $P_{\text{base}}$,
* **feedback from symptoms** through a slowly evolving average $m_{\text{slow}}$,
* **Gaussian noise**.

Discrete-time dynamics (Euler–Maruyama):

$$
m_{\text{slow}, t+1} = m_{\text{slow}, t} + \lambda_m (m_t - m_{\text{slow}, t}),
$$

$$
  P_{t+1} = P_t + dt\left[-\kappa(P_t - P_{\text{base}}) + b(m_{\text{slow}, t} - m_*)\right] + \sigma_P\sqrt{dt},\epsilon_t.
$$

```{r}
#| label: slowfast
#| include: true

simulate_slowfast_cw01 <- function(
  T_steps = 4000,
  dt      = 0.02,
  P_base  = 0,
  P0      = 0,
  m0      = 0.1,     # initial mean in [0,1]
  kappa   = 0.2,
  b       = 0.3,
  m_star  = 0.3,
  sigmaP  = 0.03,
  lambda_m = 0.003,
  # fast layer params
  n_nodes = 100,
  sweeps  = 30,
  beta    = 1.0,
  J       = 1.0,
  h0      = 0.0,
  gammaP  = 0.5,
  seed    = 1
){
  set.seed(seed)
  
  P      <- numeric(T_steps)
  m_fast <- numeric(T_steps)
  m_slow <- numeric(T_steps)
  
  P[1]      <- P0
  m_fast[1] <- m0
  m_slow[1] <- m0
  
  # initial spins in {0,1}
  s <- rbinom(n_nodes, 1, m0)
  
  for (t in 1:(T_steps - 1)) {
    # --- fast update ---
    s <- ising_fast_step01(
      P      = P[t],
      s      = s,
      beta   = beta,
      J      = J,
      h0     = h0,
      gammaP = gammaP,
      sweeps = sweeps
    )
    
    m_fast[t] <- mean(s)
    
    # --- slow moving average of symptoms ---
    m_slow[t+1] <- m_slow[t] + lambda_m * (m_fast[t] - m_slow[t])
    
    # --- slow precarity drift ---
    drift <- -kappa * (P[t] - P_base) + b * (m_slow[t] - m_star)
    P[t+1] <- P[t] + dt * drift + rnorm(1, 0, sigmaP * sqrt(dt))
  }
  
  # final fast snapshot for plotting
  m_fast[T_steps] <- m_fast[T_steps - 1]
  
  tibble(
    t      = 1:T_steps,
    P      = P,
    m      = m_fast,
    m_slow = m_slow
  )
}
```

---

# 4. Mean-field bifurcation of the 0/1 Curie–Weiss layer

For 0/1 spins, the mean-field fixed point satisfies:

$$
m = \sigma\left(\beta\left[J(m - \tfrac{1}{2}) + h_0 + \gamma_P P\right]\right).
$$

We solve this numerically and classify stability via

$$
\frac{df}{dm}
= \beta J \sigma(H)\big(1 - \sigma(H)\big) - 1,
\quad H = J(m - 1/2) + h_0 + \gamma_P P.
$$

```{r}
#| label: bifurcation
#| include: true

cw01_fixed_points <- function(P, beta, J, h0, gammaP){
  f <- function(m) inv_logit(beta * (J*(m - 0.5) + h0 + gammaP*P)) - m
  
  df <- function(m){
    p <- inv_logit(beta * (J*(m - 0.5) + h0 + gammaP*P))
    beta * J * p * (1 - p) - 1
  }
  
  grid <- seq(0, 1, length.out = 801)
  vals <- f(grid)
  idx  <- which(diff(sign(vals)) != 0)
  
  roots <- c()
  for (k in idx) {
    r <- try(uniroot(f, c(grid[k], grid[k+1]))$root, silent = TRUE)
    if (!inherits(r, "try-error")) roots <- c(roots, r)
  }
  roots <- sort(unique(round(roots, 6)))
  if (!length(roots)) return(NULL)
  
  tibble(
    P      = P,
    m      = roots,
    stable = df(roots) < 0
  )
}

cw01_bifurcation_curve <- function(P_grid, beta, J, h0, gammaP){
  map_dfr(P_grid, ~ cw01_fixed_points(.x, beta, J, h0, gammaP))
}

## Parameters for strongly nonlinear fast layer (used in B, C)
beta_BC   <- 2.0
J_BC      <- 3.0    # beta*J = 6 > 4 → bistability possible
h0_BC     <- 0.0
gammaP_BC <- 1.0

P_grid_BC <- seq(-2, 2, length.out = 400)

bf_BC <- cw01_bifurcation_curve(P_grid_BC, beta_BC, J_BC, h0_BC, gammaP_BC)

stab_by_P <- bf_BC %>%
  filter(stable) %>%
  group_by(P) %>%
  summarise(n_stable = n(),
            m_low  = min(m),
            m_high = max(m),
            .groups = "drop")

bistable_window <- stab_by_P %>%
  filter(n_stable == 2)

if (nrow(bistable_window) > 0) {
  P_low_fold  <- min(bistable_window$P)
  P_high_fold <- max(bistable_window$P)
} else {
  # fallback values so script never breaks
  P_low_fold  <- -0.5
  P_high_fold <-  0.5
}

cat(
  "Bistable P-window (fast layer): [",
  round(P_low_fold, 2), ", ",
  round(P_high_fold, 2), "]\n", sep = ""
)

ggplot() +
  {
    if (nrow(bistable_window) > 0)
      geom_ribbon(
        data = bistable_window,
        aes(x = P, ymin = m_low, ymax = m_high),
        fill = "grey85", alpha = 0.6
      )
  } +
  geom_line(
    data = bf_BC %>% filter(!stable),
    aes(P, m),
    linetype = "dashed", colour = "grey60"
  ) +
  geom_line(
    data = bf_BC %>% filter(stable, m < 0.5),
    aes(P, m, colour = "Low branch"),
    linewidth = 1
  ) +
  geom_line(
    data = bf_BC %>% filter(stable, m >= 0.5),
    aes(P, m, colour = "High branch"),
    linewidth = 1
  ) +
  scale_color_manual(values = c("Low branch" = "#4E79A7",
                                "High branch" = "#E15759"),
                     name = NULL) +
  labs(
    title = "Bifurcation diagram of fast 0/1 Curie–Weiss layer",
    subtitle = "Grey band: bistable P-region (two stable m-values)",
    x = "Precarity P",
    y = "Mean symptom activation m"
  ) +
  theme_minimal(base_size = 13)
```

---

# 5. Parallel helper: simulate low vs high groups

```{r}
#| label: parallel-helper
#| include: true

run_scenario_parallel <- function(
  par_fast,
  par_slow,
  P_bases,
  labels,
  T_steps,
  seeds
){
  stopifnot(length(P_bases) == length(labels),
            length(labels)  == length(seeds))
  
  param_list <- pmap(
    list(P_base = P_bases,
         label  = labels,
         seed   = seeds),
    function(P_base, label, seed){
      list(
        args = c(
          list(
            T_steps = T_steps,
            P_base  = P_base,
            P0      = P_base,
            m0      = 0.1,
            seed    = seed
          ),
          par_fast,
          par_slow
        ),
        meta = list(group = label)
      )
    }
  )
  
  sim_list <- future_map(
    param_list,
    function(cfg){
      out <- do.call(simulate_slowfast_cw01, cfg$args)
      out$group <- cfg$meta$group
      out
    },
    .progress = FALSE
  )
  
  bind_rows(sim_list)
}
```

---

# 6. Scenario A – Weakly nonlinear, no bistability

* Parameter setting:

  - Fast layer: low coupling → βJ < 4 → no bistability.

  - Slow layer: moderate mean reversion.

  - Groups only differ in $P_{base}$.


* Mechanics: The Curie–Weiss activation function is shallow and unimodal. For any value of precarity $P_t$, the fast layer quickly relaxes to a single equilibrium $m^*(P)$.

Thus:
  
$$
    P_{\text{low}} < P_{\text{high}}
    \quad\Rightarrow\quad
    m_{\text{low}} < m_{\text{high}},
$$

but the relationship is nearly linear and weak.

```{r}
#| label: scenario-A
#| include: true

par_fast_A <- list(
  n_nodes = 100,
  sweeps  = 20,
  beta    = 1.0,
  J       = 2.5,   # beta*J = 2.5 < 4 ⇒ no bistability
  h0      = -0.2,
  gammaP  = 0.8
)

par_slow_A <- list(
  dt       = 0.02,
  kappa    = 0.30,
  b        = 0.20,
  m_star   = 0.10,
  sigmaP   = 0.10,
  lambda_m = 0.001
)

P_bases_A <- c(-0.8, 0.8)
labels_2  <- c("Low precarity", "High precarity")

sim_A <- run_scenario_parallel(
  par_fast = par_fast_A,
  par_slow = par_slow_A,
  P_bases  = P_bases_A,
  labels   = labels_2,
  T_steps  = 3000,
  seeds    = c(1, 2)
) %>%
  mutate(scenario = "A")

sim_A_long <- sim_A %>%
  select(t, P, m, group) %>%
  pivot_longer(c(P, m),
               names_to = "variable",
               values_to = "value")

ggplot(sim_A_long, aes(t, value, colour = group)) +
  geom_line(alpha = 0.8, linewidth = 0.5) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Low precarity" = "#4E79A7",
                                "High precarity" = "#E15759")) +
  labs(
    title    = "Scenario A – Weakly nonlinear (no bistability)",
    subtitle = "Groups differ mainly in P and m means; no regime switching.",
    x = "Time step", y = "Value", colour = ""
  )
```


* Interpretation:
  
  * Groups differ only in their mean levels of precarity and symptoms.
  * There is no regime switching, no nonlinear amplification, and no hysteresis.
  * This scenario behaves like a standard stress → symptoms regression model.

* What the simulation shows:
  
  * Two gently-separated trajectories.
  * Symptom levels do not jump or bifurcate.
  * This scenario is the “null model” of slow–fast coupling.
  
---

# 7. Scenario B – Bistable fast layer, P almost fixed


* Parameter setting:
  
  * Fast layer: βJ > 4 → *bistability* (two stable symptom equilibria).
  * Slow layer: *strong mean reversion* ($κ$ large) → $P$ remains near its baseline.
  * Groups differ only in $P_{\text{base}}$.

* Mechanics:  Because the fast layer is bistable, the bifurcation diagram contains two stable branches:
  
  * low-symptom attractor $m_{\text{low}}(P)$
  * high-symptom attractor $m_{\text{high}}(P)$

If a group’s baseline precarity lies:
  
* *below the bistable window* → they fall into the *low-m* basin
* *above the bistable window* → they fall into the *high-m* basin

But because $κ$ is large, each group’s $P_t$ barely moves from its baseline — they remain “locked” on their respective branches.

```{r}
#| label: scenario-B
#| include: true

par_fast_B <- list(
  n_nodes = 100,
  sweeps  = 40,
  beta    = beta_BC,
  J       = J_BC,
  h0      = h0_BC,
  gammaP  = gammaP_BC
)

par_slow_B <- list(
  dt       = 0.02,
  kappa    = 0.50,  # strong mean reversion
  b        = 0.08,  # weak symptom → P feedback
  m_star   = 0.10,
  sigmaP   = 0.10,
  lambda_m = 0.001
)

P_bases_B <- c(P_low_fold - 0.6, P_high_fold + 0.6)

sim_B <- run_scenario_parallel(
  par_fast = par_fast_B,
  par_slow = par_slow_B,
  P_bases  = P_bases_B,
  labels   = labels_2,
  T_steps  = 3000,
  seeds    = c(3, 4)
) %>%
  mutate(scenario = "B")

sim_B_long <- sim_B %>%
  select(t, P, m, group) %>%
  pivot_longer(c(P, m),
               names_to = "variable",
               values_to = "value")

ggplot(sim_B_long, aes(t, value, colour = group)) +
  geom_line(alpha = 0.8, linewidth = 0.5) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Low precarity" = "#4E79A7",
                                "High precarity" = "#E15759")) +
  labs(
    title    = "Scenario B – Bistable fast layer, P nearly fixed",
    subtitle = "Low group in low-m branch, high group in high-m branch of the same fast network.",
    x = "Time step", y = "Value", colour = ""
  )
```

* Interpretation:
  
  * Even though both groups share the same symptom network, they occupy different stable symptom equilibria.
  * The difference is *not in the network*, but in *where slow external conditions place the system on the bifurcation diagram*.

This is the tipping-point explanation of group differences: Small differences in slow socioeconomic conditions push groups into different attractors of the same symptom system.
  
* Simulation output:
  
  * Two flat symptom trajectories: one low, one high.
  * No switching or noise-driven episodes.
  * A clean demonstration of *structural separation* due to slow baselines.
  
---

# 8. Scenario C – Bistability + strong two-way feedback

We keep the same **bistable fast layer** (`par_fast_B`) but weaken mean reversion and/or increase feedback and noise so that $P_t$ can wander through the bistable region.

We consider:

* **C1**: Catastrophic drift into high-symptom regime.
* **C2**: Episodic flipping in and out of the high-symptom basin.


## 8.1 Scenario C1 – Catastrophic flip

* Parameter setting:
  
  * Same bistable fast layer as Scenario B.
  * Slow layer:
  
    * weak mean reversion ($κ$ small)
    * strong symptom feedback ($b$ large)
    * low noise

* Mechanics:
  
  Here $P_t$ can drift far from $P_{\text{base}}$. If symptoms rise slightly, strong feedback pushes precarity upward:
  
$$
    m_{\text{slow}} \uparrow \quad\Rightarrow\quad P \uparrow.
$$

This, in turn, shifts the fast layer's field upward, which increases m — creating a *positive feedback loop*.

If $P$ crosses the *upper fold* of the bifurcation diagram, the low-m state disappears and the system collapses into the high-symptom branch (“catastrophe”).

```{r}
#| label: scenario-C1
#| include: true

par_fast_C <- par_fast_B   # same bistable fast layer as Scenario B

# C1: weak pullback, strong positive feedback, moderate noise
par_slow_C1 <- list(
  dt       = 0.02,
  kappa    = 0.20,
  b        = 0.25,
  m_star   = 0.15,
  sigmaP   = 0.10,
  lambda_m = 0.001
)

P_bases_C1 <- c(P_low_fold - 0.6, (P_low_fold + P_high_fold)/2)

sim_C1 <- run_scenario_parallel(
  par_fast = par_fast_C,
  par_slow = par_slow_C1,
  P_bases  = P_bases_C1,
  labels   = labels_2,
  T_steps  = 5000,
  seeds    = c(21, 22)
) %>% mutate(scenario = "C1")

sim_C1_long <- sim_C1 %>%
  select(t, P, m, group) %>%
  pivot_longer(c(P, m),
               names_to  = "variable",
               values_to = "value")

ggplot(sim_C1_long, aes(t, value, colour = group)) +
  geom_line(alpha = 0.8, linewidth = 0.5) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Low precarity" = "#4E79A7",
                                "High precarity" = "#E15759")) +
  labs(
    title    = "Scenario C1 – Catastrophic drift into high-symptom regime",
    subtitle = "Low group remains in low-m region; high group is pushed above the upper fold and stays high-m.",
    x = "Time step", y = "Value", colour = ""
  )
```


*Interpretation:
  
  * A group with moderately elevated precarity can experience *runaway escalation* into a chronic high-symptom regime.
  * Once the upper basin is entered, symptoms remain persistently high.

  This demonstrates a *mechanistic route to chronicity* under high precarity + high feedback.

* Simulation output:

  * Low group stays stable.
  * High group shows a *sharp transition* into the high $m$ basin and remains there.
  * $P_t$ drifts upward, never returning.



---

## 8.2 Scenario C2 – Episodic flips up and down

*Parameter setting:
  
  * *(moderate κ, moderate b, higher noise → stochastic switching)*
  * Same bistable fast layer.
  * Slow layer:

    * *weaker* positive feedback,
    * *stronger mean reversion*,
    * *higher noise σₚ*.

* Mechanics:

$P$ fluctuates near the *bistable region*, occasionally being nudged above or below fold points by noise.
Crossing a fold forces the system into the opposite symptom branch:

$$
\text{cross lower fold} \Rightarrow m_{\text{high}} \to m_{\text{low}}
$$
$$
\text{cross upper fold} \Rightarrow m_{\text{low}} \to m_{\text{high}}
$$

Because $κ$ is stronger here, P often drifts back, allowing *repeated entries and exits* from the high-symptom basin.


```{r}
#| label: scenario-C2
#| include: true

# C2: episodic flipping – equilibria for low/high m sit inside the bistable window
par_slow_C2 <- list(
  dt       = 0.02,
  kappa    = 0.30,  # stronger pullback than C1
  b        = 0.15,  # weaker symptom -> P feedback
  m_star   = 0.30,  # symptoms have to be quite high to push P up a lot
  sigmaP   = 0.15,  # fairly big noise so P can cross folds
  lambda_m = 0.001
)

P_bases_C2 <- c(P_low_fold - 0.8,
                (P_low_fold + P_high_fold)/2)

sim_C2 <- run_scenario_parallel(
  par_fast = par_fast_C,
  par_slow = par_slow_C2,
  P_bases  = P_bases_C2,
  labels   = labels_2,
  T_steps  = 6000,
  seeds    = c(31, 32)
) %>%
  mutate(scenario = "C2")

sim_C2_long <- sim_C2 %>%
  select(t, P, m, group) %>%
  pivot_longer(c(P, m),
               names_to  = "variable",
               values_to = "value")

ggplot(sim_C2_long, aes(t, value, colour = group)) +
  geom_line(alpha = 0.8, linewidth = 0.5) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Low precarity"  = "#4E79A7",
                                "High precarity" = "#E15759")) +
  labs(
    title    = "Scenario C2 – Episodic flipping in and out of high-symptom regime",
    subtitle = "High group hovers in the bistable window and occasionally flips up to high m and back down.",
    x = "Time step", y = "Value", colour = ""
  )
```

---

# 9. Multi-replicate simulations (A, B, C1, C2)

Now we run *multiple stochastic replicates* per scenario and group to see how
robust the qualitative behaviour is.

```{r}
#| label: scenario-config

scenario_config <- function(scenario){
  switch(
    scenario,
    "A" = list(
      par_fast = par_fast_A,
      par_slow = par_slow_A,
      P_bases  = P_bases_A,
      labels   = labels_2,
      T_steps  = 6000
    ),
    "B" = list(
      par_fast = par_fast_B,
      par_slow = par_slow_B,
      P_bases  = P_bases_B,
      labels   = labels_2,
      T_steps  = 6000
    ),
    "C1" = list(
      par_fast = par_fast_C,
      par_slow = par_slow_C1,
      P_bases  = P_bases_C1,
      labels   = labels_2,
      T_steps  = 6000
    ),
    "C2" = list(
      par_fast = par_fast_C,
      par_slow = par_slow_C2,
      P_bases  = P_bases_C2,
      labels   = labels_2,
      T_steps  = 6000
    ),
    stop("Unknown scenario: ", scenario)
  )
}

simulate_replicates <- function(scenario, n_rep = 10, seed_base = 1000){
  cfg <- scenario_config(scenario)
  
  rep_list <- furrr::future_map(
    1:n_rep,
    function(r){
      seeds <- seed_base + c(2*r, 2*r + 1)
      out <- run_scenario_parallel(
        par_fast = cfg$par_fast,
        par_slow = cfg$par_slow,
        P_bases  = cfg$P_bases,
        labels   = cfg$labels,
        T_steps  = cfg$T_steps,
        seeds    = seeds
      )
      out$rep <- r
      out
    },
    .progress = FALSE
  )
  
  bind_rows(rep_list) %>% mutate(scenario = scenario)
}
```



* Interpretation:

  * The system shows episodes of high symptoms but does not remain stuck.
  * High-precarity groups show *greater volatility and more switching*.
  * This scenario is closest to describing *episodic depression / relapses* driven by social stress variability.

* Simulation output:

  * Low group stable at low m.
  * High group fluctuates, producing *intermittent spikes of high symptoms*.
  * Both m and P show characteristic “plateau–jump–plateau” dynamics.

This scenario provides an interpretable mechanistic explanation for *instability* in high-precarity populations.



```{r}
#| label: multi-rep-sims

library(furrr)
plan(multisession, workers = max(1, parallel::detectCores() - 1))

n_rep <- 30  # you can increase this later (e.g., 30) if you want

sim_reps_A  <- simulate_replicates("A",  n_rep, seed_base = 1000)
sim_reps_B  <- simulate_replicates("B",  n_rep, seed_base = 2000)
sim_reps_C1 <- simulate_replicates("C1", n_rep, seed_base = 3000)
sim_reps_C2 <- simulate_replicates("C2", n_rep, seed_base = 4000)

sim_reps_all <- bind_rows(sim_reps_A, sim_reps_B, sim_reps_C1, sim_reps_C2)
```

---

# 10. Summary plots across replicates

## 10.1 Ribbon plots (median + IQR over time)

```{r}
#| label: summary-ribbons
#| fig-width: 10
#| cache: false

library(dplyr)
library(tidyr)
library(ggplot2)

summary_long <- sim_reps_all %>%
  select(scenario, group, rep, t, P, m) %>%
  pivot_longer(c(P, m), names_to = "variable", values_to = "value") %>%
  group_by(scenario, group, variable, t) %>%
  summarise(
    median = median(value),
    lo     = quantile(value, 0.25),
    hi     = quantile(value, 0.75),
    .groups = "drop"
  )

ggplot(summary_long,
       aes(t, median, colour = group, fill = group)) +
  geom_ribbon(aes(ymin = lo, ymax = hi),
              alpha = 0.18, colour = NA) +
  geom_line(linewidth = 0.6) +
  facet_grid(variable ~ scenario, scales = "free_y") +
  scale_color_manual(values = c("Low precarity"  = "#4E79A7",
                                "High precarity" = "#E15759"),
                     name = "") +
  scale_fill_manual(values = c("Low precarity"  = "#4E79A7",
                               "High precarity" = "#E15759"),
                    guide = "none") +
  labs(
    title = "Median trajectories and IQR across replicates",
    x = "Time step", y = "Value"
  ) +
    theme(legend.position = "bottom")

```

---

## 10.2 Representative trajectories

We now choose one representative replicate per scenario × group.
We define "representative" as: mean $m$ over time closest to the group’s average across all replicates.

```{r}
#| label: representative-selection
#| include: true
#| fig-width: 10

rep_scores <- sim_reps_all %>%
  group_by(scenario, group, rep) %>%
  summarise(
    m_mean = mean(m),
    .groups = "drop"
  )

rep_targets <- rep_scores %>%
  group_by(scenario, group) %>%
  mutate(global_mean = mean(m_mean)) %>%
  slice_min(abs(m_mean - global_mean), n = 1, with_ties = FALSE) %>%
  ungroup()

rep_representative <- sim_reps_all %>%
  inner_join(rep_targets, by = c("scenario", "group", "rep"))

rep_long <- rep_representative %>%
  select(scenario, group, t, P, m) %>%
  pivot_longer(c(P, m), names_to = "variable", values_to = "value")

ggplot(rep_long, aes(t, value, colour = group)) +
  geom_line(alpha = 0.85, linewidth = 0.7) +
  facet_grid(variable ~ scenario, scales = "free_y") +
  scale_color_manual(values = c("Low precarity"  = "#4E79A7",
                                "High precarity" = "#E15759"),
                     name = "") +
  labs(
    title = "Representative trajectories per scenario and group",
    subtitle = "Rep chosen so that mean m is closest to across-replicate group mean.",
    x = "Time step", y = "Value"
  ) +
      theme(legend.position = "bottom")

```

These are good candidates for the **main time-series figures** in the paper.

---

# 11. Graphical abstract

We build a 3-panel graphical abstract:

* **(A)** Conceptual slow–fast feedback.
* **(B)** Bifurcation structure of the fast layer.
* **(C)** Mean $(P, m)$ locations of each scenario × group.

```{r}
#| label: graphical-abstract
#| fig-width: 10

## Panel A: conceptual diagram (slow–fast feedback)
panel_A <- ggplot() +
  xlim(0, 3) + ylim(0, 3) +
  annotate("text", x = 0.8, y = 2.3,
           label = "Slow precarity\nP(t)", size = 4.2, fontface = "bold") +
  annotate("text", x = 2.2, y = 2.3,
           label = "Fast symptoms\ns_i(t)", size = 4.2, fontface = "bold") +
  annotate("segment",
           x = 1.1, y = 2.1, xend = 1.9, yend = 2.1,
           arrow = arrow(length = unit(0.18, "cm")), linewidth = 0.7) +
  annotate("segment",
           x = 1.9, y = 1.9, xend = 1.1, yend = 1.9,
           arrow = arrow(length = unit(0.18, "cm")), linewidth = 0.7) +
  annotate("text", x = 1.5, y = 2.35,
           label = "P → thresholds", size = 3.2) +
  annotate("text", x = 1.5, y = 1.75,
           label = "m → precarity", size = 3.2) +
  theme_void() +
  labs(title = "(A) Slow–fast feedback loop")

## Panel B: bifurcation diagram (reuse bf_BC)
panel_B <- ggplot() +
  {
    if (nrow(bistable_window) > 0)
      geom_ribbon(
        data = bistable_window,
        aes(x = P, ymin = m_low, ymax = m_high),
        fill = "grey85", alpha = 0.6
      )
  } +
  geom_line(
    data = bf_BC %>% filter(!stable),
    aes(P, m),
    linetype = "dashed", colour = "grey60"
  ) +
  geom_line(
    data = bf_BC %>% filter(stable, m < 0.5),
    aes(P, m, colour = "Low branch"),
    linewidth = 0.9
  ) +
  geom_line(
    data = bf_BC %>% filter(stable, m >= 0.5),
    aes(P, m, colour = "High branch"),
    linewidth = 0.9
  ) +
  scale_color_manual(values = c("Low branch" = "#4E79A7",
                                "High branch" = "#E15759"),
                     name = "") +
  labs(
    title = "(B) Bifurcation of fast layer",
    x = "Precarity P", y = "Mean m"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")

## Panel C: scenario × group mean locations in (P, m)
scenario_means <- sim_reps_all %>%
  group_by(scenario, group) %>%
  summarise(
    P_mean = mean(P),
    m_mean = mean(m),
    .groups = "drop"
  )

panel_C <- ggplot(scenario_means,
                  aes(P_mean, m_mean, colour = group, shape = scenario)) +
  geom_point(size = 3) +
  scale_color_manual(values = c("Low precarity"  = "#4E79A7",
                                "High precarity" = "#E15759"),
                     name = "Group") +
  labs(
    title = "(C) Average location of each scenario×group",
    x = "Mean precarity (P)",
    y = "Mean symptoms (m)"
  ) +
  theme_minimal(base_size = 11) + 
  theme(legend.position = "bottom")
  

# Combine panels using patchwork
panel_A / (panel_B | panel_C)
```






# 12. HELIUS Application — Data-Driven Calibration of the Curie–Weiss Slow–Fast Model

This section demonstrates how group differences in HELIUS symptoms (e.g., Dutch vs non-Dutch) can be reproduced using the *data-calibrated* Curie–Weiss slow–fast model.

We proceed in four steps:

1. Clean HELIUS and construct:
   - a precarity index $zP$
   - binary symptoms for the fast layer
2. Estimate Curie–Weiss parameters from the symptom covariance:
   - effective slope $\beta J$
   - external field offset $h_0$
3. Place empirical group means onto the empirical bifurcation diagram
4. Calibrate the slow layer so that the simulated ΔP and Δm  
   match HELIUS ΔP and Δm.

Primary demonstration: **Ethnicity (Dutch vs NonDutch)**  
Other groupings (Education, Gender, Age) can be swapped with a single line.


---

## 12.1 Load & clean HELIUS

```{r}
#| label: helius-load

library(haven)
library(dplyr)
library(tidyr)
library(ggplot2)
library(purrr)
library(IsingFit)

helius <- read_sav("data/helius.sav")

precarity_vars <- c("work_sit","inc_dif","discrim","hea_lit","soc_freq")
symptom_vars   <- c("anh","dep","slp","ene","app","glt","con","mot","sui")

dat <- helius %>%
  select(
    H1_WerkSit, H1_InkHhMoeite, H1_Discr_sumscore,
    H1_SBSQ_meanscore, H1_SSQT,
    H1_WlbvRecent1,H1_WlbvRecent2,H1_WlbvRecent3,
    H1_WlbvRecent4,H1_WlbvRecent5,H1_WlbvRecent6,
    H1_WlbvRecent7,H1_WlbvRecent_89,H1_WlbvRecent10,
    H1_PHQ9_sumscore,
    H1_etniciteit, H1_geslacht, H1_Opleid, H1_lft
  ) %>%
  mutate(across(everything(), ~ na_if(.x, -9)),
         across(everything(), ~ na_if(.x, -1))) %>%
  drop_na() %>%
  mutate(across(everything(), as.numeric)) %>%
  rename(
    work_sit = H1_WerkSit,
    inc_dif  = H1_InkHhMoeite,
    discrim  = H1_Discr_sumscore,
    hea_lit  = H1_SBSQ_meanscore,
    soc_freq = H1_SSQT,
    anh = H1_WlbvRecent1, dep = H1_WlbvRecent2, slp = H1_WlbvRecent3,
    ene = H1_WlbvRecent4, app = H1_WlbvRecent5, glt = H1_WlbvRecent6,
    con = H1_WlbvRecent7, mot = H1_WlbvRecent_89, sui = H1_WlbvRecent10,
    PHQsum = H1_PHQ9_sumscore,
    ethn   = H1_etniciteit,
    gender = H1_geslacht,
    edu_cat = H1_Opleid,
    age    = H1_lft
  )
  
```

---

## 12.2 Construct precarity index + binarize symptoms

```{r}
#| label: helius-precarity

data_std <- dat %>%
  mutate(across(all_of(precarity_vars),
                ~ as.numeric(scale(.x)[,1]),
                .names = "z_{col}")) %>%
  mutate(
    precarity_index = z_work_sit + z_inc_dif + z_discrim + z_hea_lit + z_soc_freq,
    zP = as.numeric(scale(precarity_index)[,1])
  )

# Binarize PHQ items for the Curie–Weiss fast layer
data_bi <- data_std %>%
  mutate(across(all_of(symptom_vars), ~ ifelse(.x > 0, 1, 0)))
```

---

## 12.3 Group maker functions (Ethnicity default)

```{r}
#| label: helius-groups

#| label: helius-groups
#| echo: true

make_group_ethnicity <- function(df){
df %>%
mutate(
group = case_when(
ethn == 1        ~ "Dutch",
ethn != 1        ~ "NonDutch",
TRUE             ~ NA_character_
)
)
}

make_group_gender <- function(df){
df %>%
mutate(
group = case_when(
gender == 1 ~ "Male",
gender == 2 ~ "Female",
TRUE        ~ NA_character_
)
)
}

make_group_edu <- function(df){
df %>%
mutate(
group = case_when(
edu_cat %in% c(3,4) ~ "HighEdu",
edu_cat %in% c(1,2) ~ "LowEdu",
TRUE                ~ NA_character_
)
)
}

make_group_age <- function(df){
df %>%
mutate(
group = case_when(
age < 40  ~ "Young",
age >= 40 ~ "Old",
TRUE      ~ NA_character_
)
)
}

# helper to pick grouping by name

get_group_maker <- function(choice = c("ethnicity","gender","education","age")){
choice <- match.arg(choice)
switch(
choice,
ethnicity = make_group_ethnicity,
gender    = make_group_gender,
education = make_group_edu,
age       = make_group_age
)
}

# choose which grouping you want to use IN THIS SECTION

grouping_choice <- "ethnicity"   # <--- change to "gender","education","age" if you like
GROUP_MAKER     <- get_group_maker(grouping_choice)

```

---

## 12.4 Compute empirical group means (P, m)

(*Ethnicity used here, but the next line can be swapped easily.*)

```{r}
#| label: helius-group-summaries
#| echo: true

get_group_summary <- function(df, GROUP_MAKER){
df %>%
GROUP_MAKER() %>%
filter(!is.na(group)) %>%
group_by(group) %>%
summarise(
mean_P = mean(zP),
sd_P   = sd(zP),
mean_m = mean(rowMeans(across(all_of(symptom_vars)))),
sd_m   = sd(rowMeans(across(all_of(symptom_vars)))),
n      = n(),
.groups = "drop"
)
}

emp_groups <- get_group_summary(data_bi, GROUP_MAKER)
emp_groups
```

```{r}
#| code-fold: false

# empirical contrasts

Dslow_emp <- abs(diff(emp_groups$mean_P))
Dfast_emp <- abs(diff(emp_groups$mean_m))
ratio_emp <- Dfast_emp / Dslow_emp

Dslow_emp; Dfast_emp; ratio_emp

```

---

## 12.5 Estimate fast-layer Curie–Weiss parameters from HELIUS

We treat HELIUS symptom correlations as arising from a mean-field CW model:

$$
m = \sigma\big(\beta[J(m - 1/2) + h_0 + \gamma_P P]\big)
$$

Data identify:

* **slope near mean** → $\beta J$
* **intercept** → $h_0^\text{eff}$

### 12.5.1 Compute empirical correlation matrix

```{r}
#| label: cw-calibration
#| code-fold: false

# 1) aggregate HELIUS symptoms per participant

sym_mat <- data_bi %>% select(all_of(symptom_vars))
sum_sym <- sym_mat %>% transmute(sum_symptoms = rowSums(across(everything())))
df_glm <- data_bi %>%
select(zP) %>%
bind_cols(sum_sym)

# 2) binomial GLM: 9 trials per person (9 PHQ items)

fit_glm <- glm(
cbind(sum_symptoms, 9 - sum_symptoms) ~ zP,
data   = df_glm,
family = binomial()
)

summary(fit_glm)

alpha_hat <- coef(fit_glm)[1]
gamma_hat <- coef(fit_glm)[2]

# 3) empirical mean m across all participants

m_bar <- mean(rowMeans(sym_mat))

# 4) Use structural β,J from Scenario B/C (bistable CW)

beta_H <- beta_BC   # from earlier section (e.g., 2.0)
J_H    <- J_BC      # e.g., 3.0

# 5) Solve for h0 such that at P=0, m ≈ m_bar

# logit(m_bar) = β[J(m_bar - 0.5) + h0]   when P=0

# ⇒ h0 = 1/β * logit(m_bar) - J(m_bar - 0.5)

h0_H <- (qlogis(m_bar) / beta_H) - J_H * (m_bar - 0.5)

# 6) γP from GLM slope: γ_hat ≈ β γP  ⇒ γP = γ_hat / β

gammaP_H <- gamma_hat / beta_H

c(beta_H = beta_H, J_H = J_H, h0_H = h0_H, gammaP_H = gammaP_H)

```


<!-- ### **12.5.2 Estimate βJ from correlation** -->

<!-- For a CW model with 0/1 spins, the mean-field susceptibility gives: -->

<!-- [ -->
<!-- \bar r \approx \frac{\beta J}{1 + \beta J} \implies -->
<!-- \beta J \approx \frac{\bar r}{1-\bar r}. -->
<!-- ] -->

<!-- ```{r} -->
<!-- #| label: cw-estimate-betaJ -->

<!-- betaJ_hat <- mean_corr / (1 - mean_corr) -->
<!-- betaJ_hat -->
<!-- ``` -->

<!-- --- -->

<!-- ### **12.5.3 Estimate external field h₀** -->

<!-- We invert the fixed-point equation: -->

<!-- [ -->
<!-- m = \sigma(\beta[J(m-1/2) + h_0]) -->
<!-- ] -->

<!-- Solve for (h_0): -->

<!-- ```{r} -->
<!-- #| label: cw-estimate-h0 -->

<!-- m_hat <- mean(colMeans(sym_mat))     # empirical mean of 0/1 symptoms -->
<!-- beta_hat <- 1                        # we treat beta=1 and J=betaJ_hat -->

<!-- J_hat <- betaJ_hat -->

<!-- logit_m <- log(m_hat/(1-m_hat)) -->
<!-- h0_hat <- logit_m - J_hat*(m_hat - 0.5) -->

<!-- h0_hat -->
<!-- ``` -->

---

## 12.6 Empirical bifurcation diagram (using estimated βJ, h₀)

```{r}
#| label: cw-bifurcation-helius
#| fig-width: 8
#| fig-height: 5

P_grid_H <- seq(-5.5, 4, length.out = 400)

bf_H <- cw01_bifurcation_curve(
P_grid = P_grid_H,
beta   = beta_H,
J      = J_H,
h0     = h0_H,
gammaP = gammaP_H
)

stab_by_P_H <- bf_H %>%
filter(stable) %>%
group_by(P) %>%
summarise(
n_stable = n(),
m_low    = min(m),
m_high   = max(m),
.groups  = "drop"
)

bistable_window_H <- stab_by_P_H %>%
filter(n_stable == 2)

ggplot() +
{
if (nrow(bistable_window_H) > 0)
geom_ribbon(
data = bistable_window_H,
aes(x = P, ymin = m_low, ymax = m_high),
fill = "grey85", alpha = 0.6
)
} +
geom_line(
data = bf_H %>% filter(!stable),
aes(P, m),
linetype = "dashed", colour = "grey60"
) +
geom_line(
data = bf_H %>% filter(stable, m < 0.5),
aes(P, m, colour = "Low branch"),
linewidth = 1
) +
geom_line(
data = bf_H %>% filter(stable, m >= 0.5),
aes(P, m, colour = "High branch"),
linewidth = 1
) +

# overlay HELIUS group means

geom_point(
data = emp_groups %>% rename(P = mean_P, m = mean_m),
aes(P, m, colour = group),
size = 3, shape = 16
) +
scale_color_manual(
values = c(
"Low branch" = "#4E79A7",
"High branch" = "#E15759",
# you may have 2 groups; palette extends naturally
"Dutch"      = "#4E79A7",
"NonDutch"   = "#E15759",
"Male"       = "#4E79A7",
"Female"     = "#E15759",
"HighEdu"    = "#4E79A7",
"LowEdu"     = "#E15759",
"Young"      = "#4E79A7",
"Old"        = "#E15759"
),
breaks = c("Low branch","High branch", unique(emp_groups$group)),
name   = NULL
) +
labs(
title    = "Bifurcation of calibrated CW fast layer with HELIUS groups",
subtitle = "Solid lines: stable equilibria; dashed: unstable. Points = HELIUS group means (P,m).",
x = "Precarity P (model scale ~ zP)",
y = "Mean symptom activation m"
) +
theme_minimal(base_size = 13)

```

---

## 12.7 Calibrate the slow layer using HELIUS ΔP and Δm**

We want:

$$
|\Delta m|_{\text{sim}} \approx |\Delta m|_{\text{emp}}, \quad
|\Delta P|_{\text{sim}} \approx |\Delta P|_{\text{emp}}.
$$

We will set:

* baseline $P_\text{base}$ = empirical means
* $b, \kappa, \sigma_P$ tuned to approximate empirical contrast
* $m_* =$ empirical global baseline

```{r}
#| label: helius-map-params
#| code-fold: false

# Fast layer calibrated from HELIUS:

par_fast_H <- list(
n_nodes = 100,
sweeps  = 40,
beta    = beta_H,
J       = J_H,
h0      = h0_H,
gammaP  = gammaP_H
)

# Global m baseline = empirical mean

m_star_H <- mean(rowMeans(sym_mat))

# Slow layer calibration:

# Rough moment-matching: Var(P) ~ sigmaP^2 / (2*kappa)

# Want Var(zP) ≈ 1 → choose kappa ~0.3 → sigmaP ≈ sqrt(2*0.3)

kappa_H  <- 0.30
sigmaP_H <- sqrt(2 * kappa_H)   # ~ 0.775
b_H      <- 0.20                # moderate symptom → P feedback

par_slow_H <- list(
dt       = 0.02,
kappa    = kappa_H,
b        = b_H,
m_star   = m_star_H,
sigmaP   = sigmaP_H,
lambda_m = 0.001
)

# P_base per group = empirical mean zP

P_bases_H <- emp_groups$mean_P
labels_H  <- emp_groups$group

P_bases_H; labels_H

```

---

## 12.8 Run calibrated slow–fast simulation

```{r}
#| label: helius-simulation
#| echo: true

set.seed(1)

sim_HELIUS <- run_scenario_parallel(
par_fast = par_fast_H,
par_slow = par_slow_H,
P_bases  = P_bases_H,
labels   = labels_H,
T_steps  = 5000,
seeds    = c(123, 456)   # two groups → two seeds
)

sim_H_long <- sim_HELIUS %>%
select(t, P, m, group) %>%
pivot_longer(c(P, m), names_to = "variable", values_to = "value")

sim_H_long %>% head()

```

---

## 12.9 Empirical vs simulated ΔP and Δm

```{r}
#| label: helius-vs-sim-ts
#| fig-height: 5
#| fig-width: 12

# empirical m distributions for chosen grouping

emp_ts <- data_bi %>%
GROUP_MAKER() %>%
filter(!is.na(group)) %>%
mutate(m_emp = rowMeans(across(all_of(symptom_vars))))

p_emp <- ggplot(emp_ts, aes(x = group, y = m_emp, fill = group)) +
geom_boxplot(alpha = 0.6) +
labs(
title = "Empirical HELIUS symptom levels",
y = "m (mean of 9 symptoms)"
) +
theme_minimal(base_size = 13) +
    theme(legend.position = "bottom")


p_sim <- ggplot(sim_H_long %>% filter(variable == "m"),
aes(t, value, colour = group)) +
geom_line(alpha = .6) +
labs(
title = "Simulated symptom trajectories",
y = "m"
) +
theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

p_emp | p_sim

```

<!-- Plot: -->

<!-- ```{r} -->
<!-- ggplot(tibble( -->
<!--   source=c("Empirical","Simulated"), -->
<!--   Dslow=c(Dslow_emp, Dslow_sim), -->
<!--   Dfast=c(Dfast_emp, Dfast_sim) -->
<!-- ), aes(Dslow, Dfast, label=source, color=source)) + -->
<!--   geom_point(size=4) + -->
<!--   geom_text(nudge_y=.01) + -->
<!--   geom_abline(slope=1, linetype="dashed") + -->
<!--   labs(title="Empirical vs simulated contrasts", -->
<!--        x="|Δ Precarity|", y="|Δ Symptom level|") -->
<!-- ``` -->

<!-- --- -->

## 12.10 Kernel density overlay (m and P)

```{r}
#| label: helius-vs-sim-plane
#| fig-width: 8
#| fig-height: 5

emp_points <- emp_groups %>%
rename(P = mean_P, m = mean_m)

sim_points <- sim_HELIUS %>%
group_by(group) %>%
summarise(
P = mean(P),
m = mean(m),
.groups = "drop"
)

ggplot() +
geom_point(
data = emp_points,
aes(P, m, color = group),
size = 4,
shape = 16
) +
geom_point(
data = sim_points,
aes(P, m, color = group),
size = 4,
shape = 17
) +
labs(
title    = "Empirical vs simulated (P, m) group locations",
subtitle = "Circles = HELIUS, Triangles = Calibrated model",
x        = "Precarity zP / P",
y        = "Mean symptoms m"
) +
theme_minimal(base_size = 13)

```

---

### 12.11 Empirical vs simulated |ΔP| and |Δm|
                                                    
```{r}
#| label: helius-vs-sim-diff
#| echo: true

Dslow_sim <- abs(diff(sim_points$P))
Dfast_sim <- abs(diff(sim_points$m))
ratio_sim <- Dfast_sim / Dslow_sim

df_diff <- tibble(
  type  = c("Empirical", "Simulated"),
  Dslow = c(Dslow_emp, Dslow_sim),
  Dfast = c(Dfast_emp, Dfast_sim),
  ratio = c(ratio_emp, ratio_sim)
)

df_diff

ggplot(df_diff, aes(Dslow, Dfast, label = type, color = type)) +
  geom_point(size = 4) +
  geom_text(nudge_y = 0.01) +
  geom_abline(slope = 1, linetype = "dashed") +
  xlim(0, max(df_diff$Dslow) * 1.3) +
  ylim(0, max(df_diff$Dfast) * 1.3) +
  labs(
    title = "Slow vs fast group differences: HELIUS vs calibrated model",
                                                      x     = "|Δ Precarity|",
                                                      y     = "|Δ Symptom level|"
                                                    ) +
                                                    theme_minimal(base_size = 13)
```
                                                  
---
  
### 12.12 Kernel density overlays (m and P)
  
```{r}
#| label: kde-m
#| fig-width: 8
#| fig-height: 5

emp_kde <- data_bi %>%
  GROUP_MAKER() %>%
  filter(!is.na(group)) %>%
  mutate(m_emp = rowMeans(across(all_of(symptom_vars))))

sim_kde <- sim_HELIUS %>%
  group_by(group) %>%
  mutate(m_sim = m)

ggplot() +
  geom_density(
    data = emp_kde,
    aes(m_emp, color = paste0("Emp_", group)),
    linewidth = 1.1
  ) +
  geom_density(
    data = sim_kde,
    aes(m_sim, color = paste0("Sim_", group)),
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Empirical vs simulated distributions", 
    x = "m", y = "Density"
  ) +
  theme_minimal(base_size = 13)
```

```{r}
#| label: kde-P
#| fig-width: 8
#| fig-height: 5

ggplot() +
  geom_density(
    data = emp_kde,
    aes(zP, color = paste0("Emp_", group)),
    linewidth = 1.1
  ) +
  geom_density(
    data = sim_kde,
    aes(P, color = paste0("Sim_", group)),
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Empirical vs simulated distribution",
    x = "P / zP", y = "Density"
  ) +
  theme_minimal(base_size = 13)
```

---
  
### 12.13 Network comparison 
  
This also fixes your `slice_tail` error.

```{r}
#| label: network-comparison
#| echo: true

library(tidyr)

# Empirical IsingFit network on HELIUS
W_emp <- IsingFit(
  data_bi %>% select(all_of(symptom_vars)),
  plot   = FALSE,
  family = "binomial"
)$weiadj
diag(W_emp) <- 0

# Model pseudo-network: sample Bernoulli 
# for one group (e.g., "high precarity" = second level)
target_group <- labels_H[2]

sim_group_rows <- sim_HELIUS %>%
  filter(group == target_group)

Ntail <- min(1500, nrow(sim_group_rows))

node_draws <- sim_group_rows %>%
  tail(Ntail) %>%
  mutate(prob = pmin(pmax(m, 1e-6), 1 - 1e-6)) %>%
  mutate(nodes = map(prob, ~ rbinom(9, 1, .x))) %>%
  select(-prob) %>%
  unnest_wider(nodes, names_sep = "_")

W_model <- cor(node_draws %>% select(starts_with("nodes_")))
diag(W_model) <- 0

df_net <- tibble(
  emp = as.vector(W_emp),
  mod = as.vector(W_model)
) %>%
  filter(!is.na(emp), !is.na(mod))

ggplot(df_net, aes(emp, mod)) +
  geom_point(alpha = .6) +
  geom_smooth(method = "lm", se = FALSE, colour = "black") +
  labs(
    title = "Empirical IsingFit vs model pseudo-network",
    x     = "Empirical edge weight (IsingFit)",
    y     = "Model-implied edge correlation"
  ) +
  theme_minimal(base_size = 13)
```
                                                  