---
title: "Slow–Fast Hybrid Model: 0/1 Curie–Weiss Symptoms & Precarity"
author: "Kyuri Park"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: cosmo
execute:
  warning: false
  message: false
---


```{r}
#| label: setup
#| include: true

## Core libraries
library(dplyr)
library(tidyr)
library(purrr)
library(tibble)
library(ggplot2)

## Parallelisation
library(future)
library(furrr)

theme_set(theme_minimal(base_size = 13))

## Logistic helper
inv_logit <- function(x) 1/(1 + exp(-x))

## Parallel backend
plan(multisession, workers = max(1, parallel::detectCores() - 1))
```

---

# 1. Motivation

Mental health symptoms fluctuate on fast time scales (hours–days), while
socio-economic pressures such as financial precarity, discrimination, or
workload change more slowly (months–years). Standard symptom-network models:

* treat symptoms as a **closed system**,
* interpret group differences in network structure as **internal differences**.

Here we explicitly couple:

* a **fast 0/1 Curie–Weiss symptom layer** (binary symptoms with mean-field interaction),
* a **slow precarity variable** (P_t) that drifts gradually,
* **feedback** from symptoms back into precarity.

We show how **group differences in symptom trajectories** can arise purely from:

* different **precarity baselines** and
* different **feedback strengths**,

with **identical** fast-layer parameters.

We examine:

* **Scenario A**: Weakly nonlinear fast layer (no bistability).
* **Scenario B**: Strongly nonlinear fast layer with bistability; precarity almost fixed around different baselines.
* **Scenario C1**: Same bistable fast layer + strong two-way feedback → catastrophic drift into high-symptom regime.
* **Scenario C2**: Same bistable fast layer + moderate feedback + noise → episodic flipping in and out of high-symptom regime.

---

# 2. Fast layer: 0/1 Curie–Weiss symptoms

We use a **Curie–Weiss-type model** with spins

[
s_i \in {0, 1}.
]

The mean symptom activation is

[
m = \frac{1}{N}\sum_i s_i \in [0,1].
]

The local “field” for a node is

[
H = J(m - \tfrac{1}{2}) + h_0 + \gamma_P P.
]

We then update spins with a logistic rule:

[
\Pr(s_i = 1 \mid m, P) = \sigma\big(\beta H\big)
= \frac{1}{1 + e^{-\beta H}}.
]

For large enough (\beta J) (specifically (\beta J > 4)) this mean-field system
is **bistable** for an intermediate range of (P).

```{r}
#| label: fast-layer
#| include: true

# One fast update block for 0/1 Curie–Weiss Ising layer
# s in {0,1}; P is current precarity
ising_fast_step01 <- function(
  P,
  s,
  beta   = 1.0,
  J      = 1.0,
  h0     = 0.0,
  gammaP = 0.5,
  sweeps = 30
){
  n <- length(s)  # spins in {0,1}
  
  for (sw in seq_len(sweeps)) {
    m <- mean(s)  # mean activation in [0,1]
    
    # field for this mean-field model
    H <- J * (m - 0.5) + h0 + gammaP * P
    p <- inv_logit(beta * H)
    
    # update spins
    for (u in seq_len(n)) {
      i    <- sample.int(n, 1)
      s[i] <- rbinom(1, 1, p)
    }
  }
  s
}
```

---

# 3. Slow layer: precarity dynamics

The slow variable (P_t) evolves with:

* **mean reversion** to group-specific baseline (P_{\text{base}}),
* **feedback from symptoms** through a slowly evolving average (m_{\text{slow}}),
* **Gaussian noise**.

Discrete-time dynamics (Euler–Maruyama):

[
\begin{aligned}
m_{\text{slow}, t+1} &= m_{\text{slow}, t}

* \lambda_m (m_t - m_{\text{slow}, t}), \
  P_{t+1} &= P_t + dt,\left[-\kappa(P_t - P_{\text{base}}) + b(m_{\text{slow}, t} - m_*)\right]
* \sigma_P\sqrt{dt},\epsilon_t.
  \end{aligned}
  ]

```{r}
#| label: slowfast
#| include: true

simulate_slowfast_cw01 <- function(
  T_steps = 4000,
  dt      = 0.02,
  P_base  = 0,
  P0      = 0,
  m0      = 0.1,     # initial mean in [0,1]
  kappa   = 0.2,
  b       = 0.3,
  m_star  = 0.3,
  sigmaP  = 0.03,
  lambda_m = 0.003,
  # fast layer params
  n_nodes = 100,
  sweeps  = 30,
  beta    = 1.0,
  J       = 1.0,
  h0      = 0.0,
  gammaP  = 0.5,
  seed    = 1
){
  set.seed(seed)
  
  P      <- numeric(T_steps)
  m_fast <- numeric(T_steps)
  m_slow <- numeric(T_steps)
  
  P[1]      <- P0
  m_fast[1] <- m0
  m_slow[1] <- m0
  
  # initial spins in {0,1}
  s <- rbinom(n_nodes, 1, m0)
  
  for (t in 1:(T_steps - 1)) {
    # --- fast update ---
    s <- ising_fast_step01(
      P      = P[t],
      s      = s,
      beta   = beta,
      J      = J,
      h0     = h0,
      gammaP = gammaP,
      sweeps = sweeps
    )
    
    m_fast[t] <- mean(s)
    
    # --- slow moving average of symptoms ---
    m_slow[t+1] <- m_slow[t] + lambda_m * (m_fast[t] - m_slow[t])
    
    # --- slow precarity drift ---
    drift <- -kappa * (P[t] - P_base) + b * (m_slow[t] - m_star)
    P[t+1] <- P[t] + dt * drift + rnorm(1, 0, sigmaP * sqrt(dt))
  }
  
  # final fast snapshot for plotting
  m_fast[T_steps] <- m_fast[T_steps - 1]
  
  tibble(
    t      = 1:T_steps,
    P      = P,
    m      = m_fast,
    m_slow = m_slow
  )
}
```

---

# 4. Mean-field bifurcation of the 0/1 Curie–Weiss layer

For 0/1 spins, the mean-field fixed point satisfies:

[
m = \sigma\left(\beta\left[J(m - \tfrac{1}{2}) + h_0 + \gamma_P P\right]\right).
]

We solve this numerically and classify stability via

[
\frac{df}{dm}
= \beta J ,\sigma(H)\big(1 - \sigma(H)\big) - 1,
\quad H = J(m - 1/2) + h_0 + \gamma_P P.
]

```{r}
#| label: bifurcation
#| include: true

cw01_fixed_points <- function(P, beta, J, h0, gammaP){
  f <- function(m) inv_logit(beta * (J*(m - 0.5) + h0 + gammaP*P)) - m
  
  df <- function(m){
    p <- inv_logit(beta * (J*(m - 0.5) + h0 + gammaP*P))
    beta * J * p * (1 - p) - 1
  }
  
  grid <- seq(0, 1, length.out = 801)
  vals <- f(grid)
  idx  <- which(diff(sign(vals)) != 0)
  
  roots <- c()
  for (k in idx) {
    r <- try(uniroot(f, c(grid[k], grid[k+1]))$root, silent = TRUE)
    if (!inherits(r, "try-error")) roots <- c(roots, r)
  }
  roots <- sort(unique(round(roots, 6)))
  if (!length(roots)) return(NULL)
  
  tibble(
    P      = P,
    m      = roots,
    stable = df(roots) < 0
  )
}

cw01_bifurcation_curve <- function(P_grid, beta, J, h0, gammaP){
  map_dfr(P_grid, ~ cw01_fixed_points(.x, beta, J, h0, gammaP))
}

## Parameters for strongly nonlinear fast layer (used in B, C)
beta_BC   <- 2.0
J_BC      <- 3.0    # beta*J = 6 > 4 → bistability possible
h0_BC     <- 0.0
gammaP_BC <- 1.0

P_grid_BC <- seq(-2, 2, length.out = 400)

bf_BC <- cw01_bifurcation_curve(P_grid_BC, beta_BC, J_BC, h0_BC, gammaP_BC)

stab_by_P <- bf_BC %>%
  filter(stable) %>%
  group_by(P) %>%
  summarise(n_stable = n(),
            m_low  = min(m),
            m_high = max(m),
            .groups = "drop")

bistable_window <- stab_by_P %>%
  filter(n_stable == 2)

if (nrow(bistable_window) > 0) {
  P_low_fold  <- min(bistable_window$P)
  P_high_fold <- max(bistable_window$P)
} else {
  # fallback values so script never breaks
  P_low_fold  <- -0.5
  P_high_fold <-  0.5
}

cat(
  "Bistable P-window (fast layer): [",
  round(P_low_fold, 2), ", ",
  round(P_high_fold, 2), "]\n", sep = ""
)

ggplot() +
  {
    if (nrow(bistable_window) > 0)
      geom_ribbon(
        data = bistable_window,
        aes(x = P, ymin = m_low, ymax = m_high),
        fill = "grey85", alpha = 0.6
      )
  } +
  geom_line(
    data = bf_BC %>% filter(!stable),
    aes(P, m),
    linetype = "dashed", colour = "grey60"
  ) +
  geom_line(
    data = bf_BC %>% filter(stable, m < 0.5),
    aes(P, m, colour = "Low branch"),
    linewidth = 1
  ) +
  geom_line(
    data = bf_BC %>% filter(stable, m >= 0.5),
    aes(P, m, colour = "High branch"),
    linewidth = 1
  ) +
  scale_color_manual(values = c("Low branch" = "#4E79A7",
                                "High branch" = "#E15759"),
                     name = NULL) +
  labs(
    title = "Bifurcation diagram of fast 0/1 Curie–Weiss layer",
    subtitle = "Grey band: bistable P-region (two stable m-values)",
    x = "Precarity P",
    y = "Mean symptom activation m"
  ) +
  theme_minimal(base_size = 13)
```

---

# 5. Parallel helper: simulate low vs high groups

We’ll repeatedly call `simulate_slowfast_cw01()` with different baselines
and seeds, running **in parallel** via `furrr::future_map`.

```{r}
#| label: parallel-helper
#| include: true

run_scenario_parallel <- function(
  par_fast,
  par_slow,
  P_bases,
  labels,
  T_steps,
  seeds
){
  stopifnot(length(P_bases) == length(labels),
            length(labels)  == length(seeds))
  
  param_list <- pmap(
    list(P_base = P_bases,
         label  = labels,
         seed   = seeds),
    function(P_base, label, seed){
      list(
        args = c(
          list(
            T_steps = T_steps,
            P_base  = P_base,
            P0      = P_base,
            m0      = 0.1,
            seed    = seed
          ),
          par_fast,
          par_slow
        ),
        meta = list(group = label)
      )
    }
  )
  
  sim_list <- future_map(
    param_list,
    function(cfg){
      out <- do.call(simulate_slowfast_cw01, cfg$args)
      out$group <- cfg$meta$group
      out
    },
    .progress = FALSE
  )
  
  bind_rows(sim_list)
}
```

---

# 6. Scenario A – Weakly nonlinear, no bistability

Here the fast layer has **weak coupling** ((\beta J < 4)), so there is a
single attractor for all (P). Groups differ in baseline precarity.

```{r}
#| label: scenario-A
#| include: true

par_fast_A <- list(
  n_nodes = 100,
  sweeps  = 20,
  beta    = 1.0,
  J       = 2.5,   # beta*J = 2.5 < 4 ⇒ no bistability
  h0      = -0.2,
  gammaP  = 0.8
)

par_slow_A <- list(
  dt       = 0.02,
  kappa    = 0.30,
  b        = 0.20,
  m_star   = 0.10,
  sigmaP   = 0.10,
  lambda_m = 0.001
)

sim_A <- run_scenario_parallel(
  par_fast = par_fast_A,
  par_slow = par_slow_A,
  P_bases  = c(-0.8, 0.8),
  labels   = c("Low precarity", "High precarity"),
  T_steps  = 3000,
  seeds    = c(1, 2)
) %>%
  mutate(scenario = "A")

sim_A_long <- sim_A %>%
  select(t, P, m, group) %>%
  pivot_longer(c(P, m),
               names_to = "variable",
               values_to = "value")

ggplot(sim_A_long, aes(t, value, colour = group)) +
  geom_line(alpha = 0.8, linewidth = 0.5) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Low precarity" = "#4E79A7",
                                "High precarity" = "#E15759")) +
  labs(
    title    = "Scenario A – Weakly nonlinear (no bistability)",
    subtitle = "Groups differ mainly in P and m means; no regime switching.",
    x = "Time step", y = "Value", colour = ""
  )
```

---

# 7. Scenario B – Bistable fast layer, P almost fixed

Now use the **bistable** fast-layer parameters from Section 4. Precarity is
tightly mean-reverting, so each group stays near its baseline (P_{\text{base}}).

* Low group baseline below bistable window → low-symptom branch.
* High group baseline above bistable window → high-symptom branch.

```{r}
#| label: scenario-B
#| include: true

par_fast_B <- list(
  n_nodes = 100,
  sweeps  = 40,
  beta    = beta_BC,
  J       = J_BC,
  h0      = h0_BC,
  gammaP  = gammaP_BC
)

par_slow_B <- list(
  dt       = 0.02,
  kappa    = 0.50,  # strong mean reversion
  b        = 0.08,  # weak symptom → P feedback
  m_star   = 0.10,
  sigmaP   = 0.10,
  lambda_m = 0.001
)

sim_B <- run_scenario_parallel(
  par_fast = par_fast_B,
  par_slow = par_slow_B,
  P_bases  = c(P_low_fold - 0.6, P_high_fold + 0.6),
  labels   = c("Low precarity", "High precarity"),
  T_steps  = 3000,
  seeds    = c(3, 4)
) %>%
  mutate(scenario = "B")

sim_B_long <- sim_B %>%
  select(t, P, m, group) %>%
  pivot_longer(c(P, m),
               names_to = "variable",
               values_to = "value")

ggplot(sim_B_long, aes(t, value, colour = group)) +
  geom_line(alpha = 0.8, linewidth = 0.5) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Low precarity" = "#4E79A7",
                                "High precarity" = "#E15759")) +
  labs(
    title    = "Scenario B – Bistable fast layer, P nearly fixed",
    subtitle = "Low group in low-m branch, high group in high-m branch of the same fast network.",
    x = "Time step", y = "Value", colour = ""
  )
```

---

# 8. Scenario C – Bistability + strong two-way feedback

We keep the same **bistable fast layer** (par_fast_B) but weaken mean
reversion and/or increase feedback and noise so that (P_t) can wander through
the bistable region.

We consider:

* **C1**: Catastrophic drift into high-symptom regime.
* **C2**: Episodic flipping in and out of the high-symptom basin.

## 8.1 Scenario C1 – Catastrophic flip

```{r}
#| label: scenario-C1
#| include: true

# --- replace your current par_slow_C1 + sim_C1 block with this ---

par_fast_C <- par_fast_B   # same bistable fast layer as Scenario B

# C1: weak pullback, strong positive feedback, low noise
par_slow_C1 <- list(
  dt       = 0.02,
  kappa    = 0.20,
  b        = 0.25,
  m_star   = 0.15,
  sigmaP   = 0.10,
  lambda_m = 0.001
)

sim_C1 <- run_scenario_parallel(
  par_fast = par_fast_C,
  par_slow = par_slow_C1,
  P_bases  = c(P_low_fold - 0.6, (P_low_fold + P_high_fold)/2),
  labels   = c("Low precarity", "High precarity"),
  T_steps  = 5000,
  seeds    = c(21, 22)
) %>% mutate(scenario = "C1")


sim_C1_long <- sim_C1 %>%
  select(t, P, m, group) %>%
  pivot_longer(c(P, m),
               names_to  = "variable",
               values_to = "value")

ggplot(sim_C1_long, aes(t, value, colour = group)) +
  geom_line(alpha = 0.8, linewidth = 0.5) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Low precarity" = "#4E79A7",
                                "High precarity" = "#E15759")) +
  labs(
    title    = "Scenario C1 – Catastrophic drift into high-symptom regime",
    subtitle = "Low group remains in low-m region; high group is pushed above the upper fold and stays high-m.",
    x = "Time step", y = "Value", colour = ""
  )

```

---

## 8.2 Scenario C2 – Episodic flips up and down

```{r}
#| label: scenario-C2
#| include: true

# --- REPLACE your current Scenario C2 block with this ---

# C2: episodic flipping – equilibria for low/high m sit inside the bistable window
par_slow_C2 <- list(
  dt       = 0.02,
  kappa    = 0.30,  # stronger pullback than before
  b        = 0.15,  # weaker symptom -> P feedback
  m_star   = 0.30,  # symptoms have to be quite high to push P up a lot
  sigmaP   = 0.15,  # fairly big noise so P can cross folds
  lambda_m = 0.001
)

# For the low group: baseline safely below the window
# For the high group: baseline near the middle of the window
sim_C2 <- run_scenario_parallel(
  par_fast = par_fast_C,          # same bistable fast layer as B/C1
  par_slow = par_slow_C2,
  P_bases  = c(P_low_fold - 0.8,           # low precarity group
               (P_low_fold + P_high_fold)/2),  # high precarity group
  labels   = c("Low precarity", "High precarity"),
  T_steps  = 6000,
  seeds    = c(31, 32)
) %>%
  mutate(scenario = "C2")

sim_C2_long <- sim_C2 %>%
  select(t, P, m, group) %>%
  pivot_longer(c(P, m),
               names_to  = "variable",
               values_to = "value")

ggplot(sim_C2_long, aes(t, value, colour = group)) +
  geom_line(alpha = 0.8, linewidth = 0.5) +
  facet_wrap(~ variable, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c("Low precarity"  = "#4E79A7",
                                "High precarity" = "#E15759")) +
  labs(
    title    = "Scenario C2 – Episodic flipping in and out of high-symptom regime",
    subtitle = "High group hovers in the bistable window and occasionally flips up to high m and back down.",
    x = "Time step", y = "Value", colour = ""
  )


```

---

# 9. Interpretation & next steps

* **Scenario A**: Fast layer is unimodal; groups differ mainly in mean levels of
  precarity and symptoms, not in qualitative dynamics.
* **Scenario B**: Fast layer is bistable; with nearly fixed P, groups occupy
  different basins of **the same Curie–Weiss network**.
* **Scenario C1**: Strong two-way coupling can create **one-way catastrophic
  transitions** into a high-symptom regime.
* **Scenario C2**: With stronger mean reversion and more noise, we get
  **episodes**: the high-precarity group flips between low- and high-symptom
  regimes as P drifts through the bistable window.

Crucially, in B/C:

> The **internal 0/1 Curie–Weiss parameters (β, J)** are identical across groups.
> Differences arise from **slow external conditions** (baseline P and feedback),
> not from different symptom couplings.

Once you’re happy with these dynamics, we can:

* build a **graphical abstract** (conceptual diagram + bifurcation + scenario
  summary), and
* add a **HELlUS section** that maps real group differences in precarity onto
  different (P_{\text{base}}), (b), and (\kappa) and shows how this
  framework explains observed symptom-network differences.
